Global code motion of congruent instructions

### 2017-11-11 cgo/paper.tex
chi symbol

So we introduce a data structure (CHI nodes) to keep track of values flowing out
of a basic block with more than once successors. We only do this for values with
multiple occurrences in the function as they are the potentially hoistable
candidates. This approach allows us to hoist instructions to a basic block with
more than two successors. The CHI node in a basic block $B$ has CHIArgs with
three entries: a) the instruction (I) CHIArg refers to, b) the basic block $Dest$
which is post-dominated by the basic block containing I, and c) the GVN of I.

@@

A CHI keeps information about `congruent' values flowing out of a basic block.
It is similar to $\phi$ \cite{ssa} but in the inverse graph, and used for
trackig out-flowing values on each edge. Although it may be suitable for other
analyses/optimizations, for \GCM{}, it is computed only for instructions with
multiple occurrences in the CFG. An illustration of CHI is shown in
Figure~\ref{fig:chi-intro}. The \GVN{} for both I1 and I2 is V, the CHI node
will save the instruction as well as the edge where the value is flowing to.
Where $Arg1$, $Arg2$ are of type CHIArg. The CHI for both I1 and I2 is inserted
in their PDF (A). For sinking we use the similar approach as \cite{ssapre} to
track values flowing into a basic block.

@@

When SSAPRE inserts $\Phi$ (expression PHIs), it does so at places where
(potentially) multiple expressions may merge. If a $\Phi$ has a missing entry,
that means the expression is partially available.  We apply a similar concept
for hoisting expression by working on inverted graph. We insert a CHI, to a
basic block with multiple successors. Then for each instruction, we register its
flow of value to its corresponding CHI which can be found in its PDF. For
conciseness we only register values with multiple occurrences. Finally, we start
walking the CFG and remove the CHI if it has a missing entry (see
Figure~\cite{fig:chi-intro2}), that means the expression is not anticipable in
the basic block having CHI. For the remaining CHIs anticipabilty of `values' is
ensured by construction.

\begin{figure}
  \includegraphics[width=0.4\textwidth]{chi-example2.pdf}
  \vspace*{-1.0cm}
\caption{I1, I2 are instructions in B, and C respectively, Arg1 = \{B, I1, V\}
  and Arg2 = \{C, I2, V\}, V is the value number for both I1, and I2. A has
  missing entry in CHI so values are not anticipable}
\label{fig:chi-intro2}
\end{figure}

@@

\subsection{Barriers}
\label{subsec:barriers}
Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. A basic-block in \LLVM{} IR is actually an
extended basic-block because there might be non-returning calls in the middle of
a basic-block. Essentially, any instruction that cannot guarantee progress is
marked as a barrier. In the absence of context, as in our current
implementation, some instructions which might be safe are still classified as
barriers, e.g., volatile loads/stores, calls with missing attributes.

@@

\subsubsection{Legality of movable instructions}
\label{subsec:legality}
Since the equality of movable candidates is purely based on the value numbers,
we also need to establish if hoisting them or sinking them would be legal. The
completeness of CHI is a necessary condition for hoistability but not
sufficient. This is because congruence of value numbers only implies that they
compute same value if the inputs are same. So for instructions like memory
references, they may compute different values should the memory get modified
along the path. For that we need to check for intersecting side-effects along
the path.

>>>>

So we introduce a data structure ($\chi$ nodes) to keep track of values flowing out
of a basic block with more than once successors. We only do this for values with
multiple occurrences in the function as they are the potentially hoistable
candidates. This approach allows us to hoist instructions to a basic block with
more than two successors. The $\chi$ node in a basic block $B$ has CHIArgs with
three entries: a) the instruction (I) CHIArg refers to, b) the basic block $Dest$
which is post-dominated by the basic block containing I, and c) the GVN of I.

@@

A $\chi$ keeps information about `congruent' values flowing out of a basic block.
It is similar to $\phi$ \cite{ssa} but in the inverse graph, and used for
trackig out-flowing values on each edge. Although it may be suitable for other
analyses/optimizations, for \GCM{}, it is computed only for instructions with
multiple occurrences in the CFG. An illustration of $\chi$ is shown in
Figure~\ref{fig:chi-intro}. The \GVN{} for both I1 and I2 is V, the $\chi$ node
will save the instruction as well as the edge where the value is flowing to.
Where $Arg1$, $Arg2$ are of type CHIArg. The $\chi$ for both I1 and I2 is inserted
in their PDF (A). For sinking we use the similar approach as \cite{ssapre} to
track values flowing into a basic block.

@@

When SSAPRE inserts $\Phi$ (expression PHIs), it does so at places where
(potentially) multiple expressions may merge. If a $\Phi$ has a missing entry,
that means the expression is partially available.  We apply a similar concept
for hoisting expression by working on inverted graph. We insert a $\chi$, to a
basic block with multiple successors. Then for each instruction, we register its
flow of value to its corresponding $\chi$ which can be found in its PDF. For
conciseness we only register values with multiple occurrences. Finally, we start
walking the CFG and remove the $\chi$ if it has a missing entry (see
Figure~\cite{fig:chi-intro2}), that means the expression is not anticipable in
the basic block having $\chi$. For the remaining $\chi$s anticipabilty of `values'
is ensured by construction.

\begin{figure}
  \includegraphics[width=0.4\textwidth]{chi-example2.pdf}
  \vspace*{-1.0cm}
\caption{I1, I2 are instructions in B, and C respectively, Arg1 = \{B, I1, V\}
  and Arg2 = \{C, I2, V\}, V is the value number for both I1, and I2. A has
  missing entry in $\chi$ so values are not anticipable}
\label{fig:chi-intro2}
\end{figure}

@@

\subsection{Barriers}
\label{subsec:barriers}
Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. A basic-block in \LLVM{} IR is actually an
extended basic-block because there might be non-returning calls in the middle of
a basic-block. Essentially, any instruction that cannot guarantee progress is
marked as a barrier. In the absence of context, as in our current
implementation, some instructions which might be safe are still classified as
barriers, e.g., calls with missing attributes.

@@

\subsubsection{Legality of movable instructions}
\label{subsec:legality}
Since the equality of movable candidates is purely based on the value numbers,
we also need to establish if hoisting them or sinking them would be legal. The
completeness of $\chi$ is a necessary condition for hoistability but not
sufficient. This is because congruence of value numbers only implies that they
compute same value if the inputs are same. So for instructions like memory
references, they may compute different values should the memory get modified
along the path. For that we need to check for intersecting side-effects along
the path.
